Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-27 submission/manuscript.Rmd
discussion edits and final paragraph

```
# Abstract

@@

A single genome holds the instructions to generate the myriad of cell types found within the adult organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. While many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by mutations in chromatin regulators[@gabrieleChromatinBasisNeurodevelopmental2018]. Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of liver-specific genes within adult neurons[@schaeferControlCognitionAdaptive2009]. Very few studies have investigated this cellular identity crisis in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if partial loss of brain identity contributes to neurodevelopmental impairments.

@@

We found a substantial portion of genes upregulated in _Kdm5c_-KO brain are germline genes with no known fucntions within the brain. Distinguishing the germline (i.e. cells that pass on their genetic material to the next generation) from the soma (i.e. all other cells of the body) is a key feature of multicellularity and sexual reproduction[XXX]. Previous work identifying chromatin regualtors that silence germ cell-specific transcription has predominatly focused on their repression of key marker genes, such as _Dazl_ and _Ddx4_[XXX]. To globally characterize KDM5C's role in germline gene repression, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[]. This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) the categories of germline genes directly and indirectly regulated by KDM5C.

@@

It is important to note that while we focused on KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal. For example, although primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, _Dazl_ is also expressed in na&#239;ve ESCs _in vitro_[@wellingDAZLRegulatesTet12015], the inner cell mass _in vivo_[@wellingDAZLRegulatesTet12015], and at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], but is silenced when cells transition from na&#239;ve to primed pluripotency during ESC to EpiLC differentiation[@mochizukiRepressionGermlineGenes2021]. Very recently, two screens of _Dazl_-repressors in ESCs identified KDM5C as a direct regulator of _Dazl_ expression[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. Interestingly, one screen found _Kdm5c_-KO ESCs also aberrantly expressed 2-cell-specific genes, indicating KDM5C negatively regulates totipotency in ESCs[@guptaGenomewideScreenReveals2023]. We found that while KDM5C also repressess _Dazl_ expression in EpiLCs, _Kdm5c_-KO EpiLCs do not express 2-cell specific genes like _Dux_ and _Zscan4c_. Out of the four 2-cell regulators characterized in ESCs, KDM5C was the only factor whose repression of _Dazl_ was independent of _Dux_ expression[@guptaGenomewideScreenReveals2023]. Together, shis suggests the 2-cell-like state in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression, including germline genes that are also involved in pluripotency and self-renewal.

It is hypothesized distinct repressive chromatin marks are placed at germline gene promoters as the embryo transitions from na&#239;ve to primed pluripotency. Initially, germline genes are repressed by placement of hisone 2A lysine 119 monoubiquitination (H2AK119ub1) and histone 3 lysine 9 trimethylation (H3K9me3) in ESCs and then gain _de novo_ DNA CpG methylation (CpGme) in EpiLCs[@velascoDnmt3bRecruitmentE2F62010; @karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. KDM5C may be instead required to remove an active mark, histone 3 lysine 4 trimethylation (H3K4me3), since H3K4me3 can impede _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. This is supported by previous work in the _Kdm5C_-KO adult hippocampus, which found an increase in H3K4me3 around the transcription start site (TSS) of germline DEGs and loss of CpGme at at least two germline promoters[@scandagliaLossKdm5cCauses2017]. However, KDM5C's role in embryonic germline gene silencing is still unclear, ask KDM5C's catalytic activity was recently found to be dispensible for silencing DAZL in ESCs[@guptaGenomewideScreenReveals2023]. In this study, we observed a global increase in H3K4me3 around the TSS of germline-enriched gene in the _Kdm5c_-KO amygdala and an increase in H3K4me2 in _Kdm5c_-KO EpiLCs. We found KDM5C's expression is dynamically regulated during ESC to EpiLC differentiation and loss of KDM5C leads to impaired placement of CpGme in extended EpiLCs. Altogether, this suggests KDM5C is necessary during the transition from na&#239;ve to primed pluripotency to promote the initial placement of CpGme at CpG islands for the long-term silencing of germline genes.

- Talk about impact - Other germline gene chromatin regulators, NDDs, cancer. Maybe talk about myt1l
- include cancer in there somewhere (Somatic misexpression of germline genes has been implicated in many cancers.)

In summary, we demonstrate loss of the X-linked intellectual disability gene KDM5C leads to widespread dysregulation of brain tissue identity, including misexpression of germline genes in the somatic brain. In EpiLCs, KDM5C directly represses key drivers of germline identity like _Dazl_, by promoting intial CpG methylation placement in the post-implantation embryo. However, germline genes can also become ectopically expressesed in _Kdm5c_-KO cells independent of direct KDM5C regulation, including the meiotic transcription factor _Stra8_. These ectopic germline developmental programs can, to some extent, mimic typical germ cell development, resulting in aberrant transcription early developmental and meiotic genes in _Kdm5c_-KO EpiLCs and late-stage spermatogenesis genes in the _Kdm5c_-KO brain. Altogether, these results define KDM5C's role in the demarcation between soma and gemrline identity and offers novel insight into how this dysregulation of tissue identity changes over the course of development. Additionally, this study provides the mechanistic foundation required to ultimately investigate the impact of aberrant germline identity upon neurodevelopment.

@@

# Acknowledgements

>>>>

```
Katherine M. Bonefas, Ilakkiya Venkatachalam, and Shigeki Iwase. 
# Abstract

@@

A single genome holds the instructions to generate the myriad of cell types found within the adult organism. This is, in part, accomplished by chromatin regulators that can either promote or impede lineage-specific gene expression through DNA and histone modifications[@strahlLanguageCovalentHistone2000; @jenuweinTranslatingHistoneCode2001]. While many chromatin regulators were initially identified for their roles in shaping cellular and tissue identity[@lewisGeneComplexControlling1978a; @kennisonDosagedependentModifiersPolycomb1988a; @kassisPolycombTrithoraxGroup2017], recent advancements in next generation sequencing unexpectedly revealed many neurodevelopmental disorders (NDDs) are caused by mutations in chromatin regulators[@gabrieleChromatinBasisNeurodevelopmental2018]. Several studies have suggested this connection between chromatin regulators and neurodevelopment is due to their regulation of brain-specific genes, such as orchestrating transcriptional programs for synaptic maturation[@zhouBrainspecificPhosphorylationMeCP22006] and transitioning between neuronal and glial fates during neural precursor differentiation[@hirabayashiPolycombLimitsNeurogenic2009]. However, loss of some chromatin regulators can also result in the ectopic transcription of tissue-specific genes outside of their target environment, such as the misexpression of heart[@mallMyt1lSafeguardsNeuronal2017;@weigelMYT1LHaploinsufficiencyHuman2023] and liver-specific[@schaeferControlCognitionAdaptive2009] genes within adult neurons. Very few studies have investigated this cellular identity crisis in chromatin-linked NDDs[@schaeferControlCognitionAdaptive2009; @scandagliaLossKdm5cCauses2017] and it is currently unknown if partial loss of brain identity contributes to neurodevelopmental impairments.

@@

We found a substantial portion of genes upregulated in _Kdm5c_-KO brain are germline genes with no known fucntions within the brain. Distinguishing the germline (i.e. cells that pass on their genetic material to the next generation) from the soma (i.e. all other cells of the body) is a key feature of multicellularity and sexual reproduction. Previous work characterizing chromatin regualtors that silence germ cell-specific transcription has predominatly focused on their repression of key marker genes in embryonic stem cells (ESCs), such as _Dazl_ and _Ddx4_[@endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. To globally characterize KDM5C's role in germline gene repression throughout life, we curated a list of mouse germline-enriched genes using publically available germ cell-depleted RNA-seq datasets from Kit^W/Wv^ mice[@muellerIndependentSpecializationHuman2013; @sohGeneRegulatoryProgram2015]. This resource enabled us to identify 1) the extent of germline gene dysregulation in _Kdm5c_-KO cells, 2) the types of germline genes misexpressed at different developmental time points, and 3) the categories of germline genes directly and indirectly regulated by KDM5C.

@@

It is important to note that while we focused on KDM5C's regulation of germline genes, some germline-enriched genes are also expressed in na&#239;ve ESCs and at the 2-cell stage due to their roles in pluripotency and self-renewal. For example, _Dazl_ is primarily known for committing PGCs to the germline fate and regulating the translation of germline-specific RNAs, but is also expressed at the 2-cell stage[@yangChemicalinducedChromatinRemodeling2022], in na&#239;ve ESCs[@wellingDAZLRegulatesTet12015], and in the inner cell mass [@wellingDAZLRegulatesTet12015]. Very recently, two screens of _Dazl_-repressors identified KDM5C as a direct regulator in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023]. Interestingly, one study found _Kdm5c_-KO ESCs also aberrantly expressed 2-cell-specific genes, indicating KDM5C negatively regulates totipotency[@guptaGenomewideScreenReveals2023]. However, out of the four regulators characterized, KDM5C was the only factor whose repression of _Dazl_ was independent of the 2-cell specific transcription factor _Dux_[@guptaGenomewideScreenReveals2023]. Although expressed at the 2-cell stage and in ESCs, _Dazl_ is typically silenced when ESCs differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021]. We found KDM5C also repressess _Dazl_ in EpiLCs, but _Kdm5c_-KO EpiLCs did not express 2-cell specific genes like _Dux_ and _Zscan4c_. Together, this suggests the 2-cell-like state in _Kdm5c_-KO ESCs is an indirect consequence of KDM5C's primary role in germline gene repression.

_Dazl_ and other germline-enriched genes are initially decommissioned in ESCs by repressive histone modifications and then _de novo_ DNA CpG methylation (CpGme) in EpiLCs[@velascoDnmt3bRecruitmentE2F62010; @karimiDNAMethylationSETDB12011; @hackettPromoterDNAMethylation2012; @tatsumiDNMTsSETDB1Function2018; @liuPolycombGroupProtein2020; @mochizukiRepressionGermlineGenes2021]. Unlike the previously characterized germline gene suppressors, KDM5C removes an active histone mark, histone 3 lysine 4 di- and trimethylation (H3K4me2/3)[@iwaseXLinkedMentalRetardation2007]. In this study, we observed a global increase in H3K4me2/3 in _Kdm5c_-KO EpiLCs and amygdala around the transcription start site (TSS) of germline-enriched genes. However, KDM5C's catalytic activity may not be required for germline gene silencing, as it was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. Although not necessary in ESCs, KDM5C's catalytic activity be required for CpGme in EpiLCs, as H3K4me3 impedes _de novo_ CpGme placement[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. This is supported by previous work in the _Kdm5C_-KO adult hippocampus, which found CpGme is significantly eroded at at least two germline promoters[@scandagliaLossKdm5cCauses2017]. To elucidate the mechanism behind KDM5C-mediated silencing of germline genes, we characterized KDM5C in ESCs and EpiLCs and found its expression is dynamically regulated during differenitation. Furthermore, KDM5C acts during the transition from na&#239;ve to primed pluripotency to promote the initial placement of CpGme at CpG islands for the long-term silencing of germline genes.

The above work provides the mechanistic foundation for KDM5C's role in germline gene repression and offers novel insight into how dysregulation in _Kdm5c_-KO tissue identity changes over the course of development. Although their impact upon neurodevelopment is currently unclear, ectopic germline transcripts are also found in models of other related neurodevelopmental disorders[@bonefasSomatogermlineTransformationChromatinlinked2021], including Immunodeficiency, Centromere instability, Facial anomalies syndrome 1 (ICF1 OMIM: #242860)[@velascoGermlineGenesHypomethylation2014; @waltonDnmt3bPrefersGerm2014], Kleefstra syndrome 1 (OMIM: #610253)[@schaeferControlCognitionAdaptive2009], and MeCP2 duplication syndrome (MDS, OMIM: #300260)[@samacoCrhOprm1Mediate2012]. Like KDM5C, the chromatin regulators underlying these conditions, DNA methyltransferase 3b (DNMT3B), H3K9me1/2 methyltransferases G9A/GLP, methyl-CpG -binding protein 2 (MECP2), primarily silence gene expression. Thus, KDM5C is among a growing cohort of chromatin-linked neurodevelopmental disorders that may have a similar underlying cause of germline versus soma dysregulation. However, further research is required to determine the ultimate impact of these germline genes and the extent to which this phenomenon occurs in humans.

@@

# Acknowledgements
- Jacob Mueller for providing insight in germline gene regulation. 
- Sundeep Kalantry for providing reagents and expertise in culturing mouse emrbyonic stem cells and epiblast like cells
- Funding sources
